Tropical mathematics for the Lambda-calculus 1

### 2023-07-29 CSL24/main.tex
Commit serale
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

\section{Tropical weighted relational semantics for high-order languagesi}\label{section3}

@@

\paragraph*{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}
\input{section3A}

\paragraph*{Allowed duplication/erasure: the $\STLC$ and the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLC+STLCO}

>>>>

\section{Topical linear/non-linear algebra}\label{section3}
\input{section3C}

\section{Tropical weighted relational semantics for high-order languages}

@@

\input{section3A}

\paragraph*{Allowed duplication/erasure: the probabilistic $\lambda$-calculus $\STLC_\oplus$}\label{sec:STLCO}

### 2023-07-29 CSL24/section2bis.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

\begin{definition}
A \emph{tropical power series} (tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$ and $I\subseteq \N$.
If $I$ is finite, $f$ is said to be a \emph{tropical polynomial}. 
\end{definition}

>>>>

\begin{definition}
A \emph{tropical power series}\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} (tPs) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
If $I$ is finite, $f$ is said to be a \emph{tropical polynomial}. 
\end{definition}

### 2023-07-29 CSL24/section3A.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

@@

>>>>

\paragraph*{Prohibited duplication/erasure: the \emph{linear} $\STLC$}\label{sec:3A}

@@

%\subsubsection{Unbounded duplications}

\paragraph*{Allowed duplication/erasure: the $\STLC$}\label{sec:STLC}

\begin{comment}

In order to interpret the full $\STLC$, we need a Cartesian closed category (CCC).
It is well-known \cite{Mellies2009} that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
%, a construction we now quickly recall.
%defined via a \emph{Lafont exponential} comonad $!$.
%Let us quickly recall the ideas behind these notions.
A SMCC is Lafont when it has finite products and it is equipped with a comonad $!$ (its \emph{Lafont exponential}) which, at level of objects, sends $X$ to an object $!X$ being the free commutative comonoid on $X$.
Such objects $!X$ represent the \emph{bang} connective of linear logic, granting infinite duplications via the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
It is well-known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ (where $\finMS{X}$ is the set of finite multi-sets on $X$) lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
Specializing [Corollary III.6, \cite{Manzo2013}] to our case, we have:

\begin{proposition}
 $\LREL$ is Lafont.
\end{proposition}

\end{comment}

As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
With no surprise $\LREL$ is Lafont~\cite[Corollary III.6]{Manzo2013} %and specialising \cite[Theorem III.7]{Manzo2013} 
w.r.t.\ the usual $!$ acting on objects by taking the set of finite multisets.
So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$, $(s\circ_! t)_{\mu,c}:=
%\begin{align}
\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.

### 2023-07-29 CSL24/section3B.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

%\subsubsection{Unbounded duplications}

\begin{comment}

In order to interpret the full $\STLC$, we need a Cartesian closed category (CCC).
It is well-known \cite{Mellies2009} that it is always possible to construct a CCC by taking the \emph{co-Kleisli} $\C C_!$ of a so-called \emph{Lafont category} $\C C$.
%, a construction we now quickly recall.
%defined via a \emph{Lafont exponential} comonad $!$.
%Let us quickly recall the ideas behind these notions.
A SMCC is Lafont when it has finite products and it is equipped with a comonad $!$ (its \emph{Lafont exponential}) which, at level of objects, sends $X$ to an object $!X$ being the free commutative comonoid on $X$.
Such objects $!X$ represent the \emph{bang} connective of linear logic, granting infinite duplications via the infinite product $X^0\otimes X\otimes X^2\otimes X^3\otimes\cdots$, each factor representing a possible number of duplications.
It is well-known that, under mild conditions satisfied by $\QREL$, one can explicit this idea via the fact that the map $X\mapsto \finMS{X}$ (where $\finMS{X}$ is the set of finite multi-sets on $X$) lifts to a functor $!:\QREL\to\QREL$ which is a Lafont-exponential comonad.
Specializing [Corollary III.6, \cite{Manzo2013}] to our case, we have:

\begin{proposition}
 $\LREL$ is Lafont.
\end{proposition}

\end{comment}

As well-known, if a SMCC is \emph{Lafont}, then one obtains a CCC from it by defining the exponential objects as $X\to Y:=!X \multimap Y$.
With no surprise $\LREL$ is Lafont (~\cite[Corollary III.6]{Manzo2013}) %and specialising \cite[Theorem III.7]{Manzo2013} 
w.r.t.\ the usual $!$ acting on objects by taking the set of finite multisets.
So the coKleisli $\LREL_!$ is CCC, i.e.\ a model of $\STLC$.
The cartesian product $\&$ of $\LREL_!$ (and of $\LREL$) is the disjoint union $+$ of sets, the \emph{evaluation} morphisms $\RM{ev}$ are matrices in $\Lawv^{!(!X\times Y)+X)\times Y}$, and the coKleisli composition of $s\in\Lawv^{!Y\times Z}$ and $t\in\Lawv^{!X\times Y}$ is the matrix $s\circ_! t\in\Lawv^{!X\times Z}$, $(s\circ_! t)_{\mu,c}:=
%\begin{align}
\inf_{n\in\N, b_1\dots,b_n\in Y, \mu = \mu_1+\cdots +\mu_n}
 \left\{s_{[b_1,\dots,b_n],c} + \sum_{i=1}^n t_{\mu_i,b_i}\right\}$.

\begin{remark}
One can see that the matrix interpreting a $\STLC$-term is always boolean (i.e.\ its coefficients are either $0$ or $\infty$) can be seen as the fact that $\STLC$ is relatively trivial, from this point of view.
This is also the case for the differential $\lambda$-calculus that we will introduce in a moment, so we state and discuss this property at that point (\autoref{prop:descrete}).
%It is instructive what its CCC-structure looks like in our tropical world.
\end{remark}

%As it is well-known, the Cartesian closed structure %of a category $\C C$  allows to define a sound interpretation $\model{\Gamma\vdash M:A}\in\HOM{\LREL_{!}}{\model \Gamma}{\model A}$ of terms as morphisms.
%In our case, we have:

@@

>>>>

@@

A precise investigation of the relation between probabilistic programs and their $\LREL_!$ interpretation is left for further studies.

### 2023-07-29 CSL24/section4.tex
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
In analogy with \cite{DanEhrh2011}, we could call them \emph{tropical entire functions}, but since, as we explain in a moment, those are generalised \emph{tropical Laurent series}, we will use the latter terminology, or \emph{tLs} for short.
The literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.

Remark that the map $(\_)^!$ is a surjective homomorphism (w.r.t.\ coKleisli composition and identity and function composition and identiy) on its image. {\color{red}In Ehrhard-Danos fanno vedere (Lemma 14) che \`e anche iniettivo. Non riesco a capire se anche a noi lo \`e, ma non vedo perch\'e non dovrebbe.}

\begin{remark}
 Since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n ! \model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is defined by $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in ! \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
Same for $\STDLC$.
Finally, the interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.
In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n ! {X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.
So, such $t^!$'s are special cases of tLs and thus, \emph{Wlog}, in the following we only consider tLs.
\end{remark}

@@

 \begin{remark}
 Our running example is indeed of shape $\varphi=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
 However, it is not the interpretation of a $\lam$-term, because its matrix $t$ is not discrete.
% Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
\end{remark}

If $\model *$ is finite, then the interpretations of $\BSTLC$-terms is a (generalised) tropical polynomial, since then the interpretation of any $\BSTLC$-type $B$ is a finite set, and thus also $!_n \model B$ is.
% 
% , i.e.~for a \emph{finite} set $\C F\subseteq \, !X$.
%%Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they 
%correspond to the tLs for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
%Actually, for us \emph{tropical polynomial} will mean a function 
%\end{remark}

@@

\begin{theorem}\label{thm:ScottCont}
 All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
 In particular, all tLs $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
\end{theorem}
\begin{proof}
 Use the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
\end{proof}

@@

The results presented in all this paper translate now into the following facts about the interpretation of higher-order programs.
Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.

>>>>

In this section we establish some topological and metric properties of such functions, this way highlighting the rich and interesting topological and metric structure of the category $\LREL_{!}$.
The literature in tropical mathematics is often recent (e.g.~\cite{Porzio2021}), and several results we prove in this section are, to our knowledge, new.

Remark that the map $(\_)^!$ is a surjective homomorphism (w.r.t.\ coKleisli composition and identity and function composition and identiy) on its image. {\color{red}In Ehrhard-Danos fanno vedere (Lemma 14) che \`e anche iniettivo. Non riesco a capire se anche a noi lo \`e, ma non vedo perch\'e non dovrebbe.}
Remark that, since $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}\in\HOM{\LREL_!}{\mathlarger{\&}_{i=1}^n \model{A_i}}{\model B}\simeq\Lawv^{\prod_{i=1}^n ! \model {A_i}\times \model B}$, the tLs $\model{(x_i:A_i)_{i=1}^n\vdash_{\STLC} M:B}^!:\Lawv^{\mathlarger{+}_{i=1}^n \model{A_i}}\to\Lawv^{\model B} \simeq \prod_{i=1}^n\Lawv^{\model{A_i}}\to\Lawv^{\model B}$ is $t^!(x^1,\dots,x^n)_b=\inf_{\mu_i\in ! \model{A_i}}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
Same for $\STDLC$.

\begin{remark}
The interpretations of $\BSTLC$-terms are matrices $t\in\HOM{\LREL}{\bigotimes_{i=1}^n !_{n_i}X_i}{Y}=\Lawv^{\prod_{i=1}^n !_{n_i}X_i\times Y}$.
In such situation, we define $t^!:\Lawv^{\mathlarger{+}_{i=1}^n X_{i}}\to\Lawv^Y \simeq \prod_{i=1}^n\Lawv^{X_{i}}\to\Lawv^Y$ as $t^!(x^1,\dots,x^n)_b:=\inf_{\mu_i\in !_{n_i} X_i}\left\{\sum_{i=1}^n\mu_i x^i+t_{(\mu_1,\dots,\mu_n),b}\right\}$.
Clearly $t^!={\widetilde t}^!$, where $\widetilde t\in\Lawv^{\prod_{i=1}^n ! {X_i}\times Y}$ is the matrix $\widetilde t_{(\mu_1,\dots,\mu_n),b}:=t_{(\mu_1,\dots,\mu_n),b}$ if $\mu_i\in !_{n_i} X_i$ for all $i$, and $:=\infty$ otherwise, which has the same support as $t$.
So, such $t^!$'s are special cases of tLs and thus we can \emph{Wlog} only consider tPs's.
\end{remark}

@@

% 
% , i.e.~for a \emph{finite} set $\C F\subseteq \, !X$.
%%Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they 
%correspond to the tLs for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}. 
%Actually, for us \emph{tropical polynomial} will mean a function 
%\end{remark}

@@

\begin{theorem}\label{thm:ScottCont}
 All monotone (w.r.t.\ pointwise order) and $\norm{\cdot}_{\infty}$-continuous functions $f:(0,\infty)^X\to (0,\infty)$ are Scott-continuous.
 In particular, all tLs $f:\Lawv^X\to\Lawv$ are Scott-continuous on $(0,\infty)^X$ w.r.t.\ the pointwise orders.
\end{theorem}
\begin{proof}
 Uses \autoref{thm:cont} plus the fact, taken from \cite{Selinger2004}, that in a normed $\R_{\geq 0}$-cone $P$, considered with its cone-order, if every bounded directed net in $P$ admits a sup, and if $(v_i)_{i\in I}$ is a directed net in $P$ with an upper bound $v\in P$, then $\exists\bigvee_{i\in I} v_i \in P$ and, if $\inf_{i\in I} \norm{v-v_i} =0$, one has $\bigvee_{i\in I} v_i = v$.
\end{proof}

@@

Let us now see what the results proved in the previous section translate into, when referred to the interpretation of higher-order programs.
Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.

### 2023-07-29 CSL24/section5.tex
Altro commit pre-cena
Commit sonnolento
La parte sulla probability ancora mi manca qualche dettaglio da chiarire, ma ci siamo quasi...

Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega_0}, Q_{\omega_0}$, which is more easily solved by 
passing to the tropical monomial/polynomial functions $\trop P_{\omega_0},\trop^{\mathrm{val}} Q_{\omega_0}$, for any fixed valuation.
For 1), by definition of $\RM{arg max}$ and because $-\log$ is stricly decreasing, we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
%\begin{equation}
%  \begin{array}{ccccc}
   \RM{arg max}_{(x,y)} P_{\omega_0} (x,y)
   %& 
   = %&
   \RM{arg min}_{(x,y)}\set{-\log P_{\omega_0} (x,y)}
   %&
   =%&
   \RM{arg min}_{(x,y)}\set{(\trop P_{\omega_0}) (-\log x,-\log y)} \label{eq:argmax}$
where this holds for any valuation (because monomials do not carry any coefficient).
Remark that $(\trop P_{\omega})( -\log x, -\log y)$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow \false_{\omega}$'', so we see that the tropicalisation allows to compute such quantities.
%  \end{array}
%\end{equation}
For 2), %the $\omega_l$ is s.t.\ $\max\limits_{\omega\in\set{ll,rll,rrr}} \, P_\omega(x,y) = P_{\omega_l}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max_{\omega\in\set{lr,rrl,rlr}} \, P_\omega(p,q) = P_{\omega_0}(p,q)$, i.e.\ $\min_{\omega\in\set{lr,rrl,rlr}} \, -\log P_\omega(p,q) = -\log P_{\omega_0}(p,q)$.
Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to asking $(\trop^{\mathrm{val}_1} Q_{\false})( -\log p, -\log q) = (\trop^{\mathrm{val}_1} P_{\omega_0})( -\log x, -\log y)\label{eq:max}$,
and this time this holds, in general, only for the trivial valuation.
Remark that, in both cases, passing through $\trop^0 $ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2lr5}), and this essentially corresponds to study negative log-probabilities. %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

@@

From this situation we notice the following important:

\begin{remark}\label{rmk:tropof01Rel}
Our toy first-order language can be interpreted both in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$ and in $\LREL$.
We do not give details now since in the rest of the paper we will interpret interesting \emph{high-order} calculi, including a probabilistic one containing our toy language.
But it is important to mention already at this point that the probabilities we just discussed above, are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
Therefore, the optimisation problems concerning the negative log-probabilities and likelihoods are already expressible by taking $\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$. 
Now, the model $\LREL$ is precisely trivial-valuation tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$, i.e.\ $\model{M}^{\LREL}=\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$.
As we remarked in \autoref{rmk:val_trop}, this corresponds to quotienting the polynomial interpreting $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$ w.r.t.\ idempotent sum.
A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
For now, just have in mind that $\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the negative log-probability of \emph{any} of the (equiprobable) \emph{most likely} reduction paths to normal form, so $\LREL$ directly expresses such optimisations problems (no need to pass through a tropicalisation).
We take these observations as incouraging, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential properties that coexist in it.
% (in $p$ and $q$, not after the substitution $q:=r-p$) 
% are extracted from it, \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
\end{remark}

>>>>

Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega_0}, Q_{\omega_0}$, which is more easily solved by 
passing to the tropical monomial/polynomial functions $\trop P_{\omega_0},\trop^{\mathrm{val}} Q_{\omega_0}$, for any fixed valuation.
For 1), by definition of $\RM{arg max}$ and because $-\log$ is stricly decreasing, we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
%\begin{equation}
%  \begin{array}{ccccc}
   \RM{arg max}_{(x,y)} P_{\omega_0} (x,y)
   %& 
   = %&
   \RM{arg min}_{(x,y)}\set{-\log P_{\omega_0} (x,y)}
   %&
   =%&
   \RM{arg min}_{(x,y)}\set{(\trop P_{\omega_0}) (-\log x,-\log y)} \label{eq:argmax}$
where this holds for any valuation (because monomials do not carry any coefficient).
Remark that $(\trop P_{\omega})( -\log x, -\log y)$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow \false_{\omega}$'', so we see that the tropicalisation allows to compute such quantities.
%  \end{array}
%\end{equation}
For 2), %the $\omega_l$ is s.t.\ $\max\limits_{\omega\in\set{ll,rll,rrr}} \, P_\omega(x,y) = P_{\omega_l}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max_{\omega\in\set{lr,rrl,rlr}} \, P_\omega(p,q) = P_{\omega_0}(p,q)$, i.e.\ $\min_{\omega\in\set{lr,rrl,rlr}} \, -\log P_\omega(p,q) = -\log P_{\omega_0}(p,q)$.
Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to asking $(\trop^{\mathrm{val}_1} Q_{\false})( -\log p, -\log q) = (\trop^{\mathrm{val}_1} P_{\omega_0})( -\log p, -\log q)\label{eq:max}$,
and this time this holds, in general, only for the trivial valuation.
Remark that, in both cases, passing through $\trop^0 $ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2lr5}), and this essentially corresponds to study negative log-probabilities. %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

@@

Our toy first-order language can be interpreted in $\LREL$.
We do not give details now since in the rest of the paper we will interpret interesting \emph{high-order} calculi, including a probabilistic one containing our toy language.
We notice the following important:

\begin{remark}\label{rmk:tropof01Rel}
$\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the minimum negative log-probability of all the reductions from $M$ to normal form, i.e.\ it coincides with the  negative log-probability of any of the (equiprobable) \emph{most} likely reductions from $M$ to normal form.
Equivalently, given $\omega_0$ a fixed reduction path from $M$ to normal form $P_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(-\log p, -\log(1-p)) 
-\log  P_{\omega_0} = \model{M}^{\LREL}_0(-\log p,-\log (1-p))$, the path $\omega_0$ is one of the \emph{most} likely one among all paths from $M$ to normal form.
Similarly, $\model{M}^{\LREL}_0(\log p,\log (1-p))$ gives the minimum log-probability of all the reductions from $M$ to normal form, i.e.\ it coincides with the log-probability of any of the (equiprobable) \emph{less} likely reductions from $M$ to normal form.
Equivalently, given $\omega_0$ a fixed reduction path from $M$ to normal form $P_{\omega_0}$, for all solution $p\in[0,1]$ to the equation $%\trop P_\omega(\log p, \log(1-p)) 
\log  P_{\omega_0} = \model{M}^{\LREL}_0(\log p,\log (1-p))$, the path $\omega_0$ is one of the \emph{less} likely one among all paths from $M$ to normal form.
% (in $p$ and $q$, not after the substitution $q:=r-p$) 
% are extracted from it, \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
\end{remark}

\begin{remark}
Our toy first-order language can be also interpreted both in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$.
But it is important to mention already at this point that the probabilities we just discussed above, are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
Therefore, the optimisation problems concerning the negative log-probabilities and likelihoods are already expressible by taking $\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$. 
Now, the model $\LREL$ is precisely trivial-valuation tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$, i.e.\ $\model{M}^{\LREL}=\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$.
As we remarked in \autoref{rmk:val_trop}, this corresponds to quotienting the polynomial interpreting $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$ w.r.t.\ idempotent sum.
A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
However, it is important to keep in mind that those properties are expressed by $\LREL$ directly with no need to pass through a tropicalsation.
\end{remark}
